---
title: "Data preparation reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data preparation reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(devPTIpack)
```

This is the **reference deep-dive** on the two data inputs that feed every PTI app:

1. A named list of administrative-boundary `sf` tibbles (saved as `admin_bounds.rds`).
2. A multi-sheet Excel metadata template describing your indicators (`mtdt.xlsx`).

If you just want a recipe, start with [Build a PTI](articles/build-pti.html). This page is for when that walkthrough refers you here for a data-shape question, or when a validator tells you something specific is wrong with your inputs.

The package's own `validate_geometries()` and `validate_metadata()` functions encode every rule on this page. If you're unsure whether your data conforms, run them — they're authoritative.

---

## 1. Boundary shapes — `admin_bounds.rds`

### 1.1 Top-level structure

`shp_dta` is a **named list of `sf` tibbles**, one element per administrative level. It is normally saved as a single `.rds` file (commonly `admin_bounds.rds`) and read with `readRDS()` at app launch.

```{r}
data(ukr_shp)

class(ukr_shp)
length(ukr_shp)
names(ukr_shp)
```

The list element names follow the package convention `admin<N>_<HumanName>`, where `<N>` is a digit 0–9 and `<HumanName>` is a single word naming the level (for example, `Oblast`, `District`, `Hexagon`). The bundled Ukraine sample uses levels 0, 1, 2, and 4 — admin3 is intentionally skipped to demonstrate that level numbers do not need to be contiguous, but each level you do include must be uniquely numbered.

### 1.2 Naming convention

| Element name      | Meaning                                                                                                |
| ----------------- | ------------------------------------------------------------------------------------------------------ |
| `admin0_Country`  | Mandatory. The country polygon. PTI apps require it even if you do not display admin0 in the UI.       |
| `admin1_Region`   | First-level subdivision (province, oblast, region, …)                                                   |
| `admin2_District` | Second-level subdivision                                                                                |
| `admin<N>_<Name>` | Pattern for any level. `<N>` is the integer level depth; `<Name>` is one human-readable word. No spaces, no colons. |

If your country uses different administrative terminology (e.g. `admin1_Wilaya`, `admin2_Commune`), use it — only the `admin<N>_` prefix matters to the package; the name suffix is for display.

### 1.3 Required columns per layer

Every `sf` tibble in the list must, at minimum, carry these columns:

| Column          | Type                  | Purpose                                                                  |
| --------------- | --------------------- | ------------------------------------------------------------------------ |
| `admin<N>Pcod`  | `character`           | Unique polygon identifier (P-code). One per row at this level.           |
| `admin<N>Name`  | `character`           | Human-readable polygon name (one per row).                               |
| `area`          | `numeric`             | Polygon area, in km². Compute after re-projecting to an appropriate UTM. |
| `geometry`      | `sfc_(MULTI)POLYGON`  | Spatial geometry. Must be `POLYGON` or `MULTIPOLYGON` only.              |

Both `admin<N>Pcod` and `admin<N>Name` use the level number of the *current* layer. For example, in `admin2_District`, those columns are `admin2Pcod` and `admin2Name`.

```{r}
str(ukr_shp$admin1_Oblast, max.level = 1)
```

### 1.4 Parent-child mapping (cascade rule)

Sub-admin layers must also carry the **parent-level P-code columns**, so the package can join up the hierarchy without re-computing keys.

The rule: a layer at depth `N` must contain the `admin<k>Pcod` column for every level `k < N` that exists elsewhere in the list. Names are not required at parent levels (only Pcods).

```{r}
intersect(c("admin0Pcod", "admin1Pcod", "admin1Name"),
          names(ukr_shp$admin1_Oblast))

intersect(c("admin0Pcod", "admin1Pcod", "admin2Pcod", "admin2Name"),
          names(ukr_shp$admin2_Rayon))
```

Schematically:

```
admin0_Country     [admin0Pcod, admin0Name]
       ▲
       │  every admin1 row carries admin0Pcod
       │
admin1_Region      [admin0Pcod, admin1Pcod, admin1Name]
       ▲
       │  every admin2 row carries admin0Pcod + admin1Pcod
       │
admin2_District    [admin0Pcod, admin1Pcod, admin2Pcod, admin2Name]
```

If a P-code value appears in `admin2_District$admin1Pcod` but does not appear in `admin1_Region$admin1Pcod`, that's an orphan child — `validate_geometries()` will flag it.

### 1.5 Geometry constraints

- **Geometry types**: `POLYGON` and `MULTIPOLYGON` only. `POINT`, `LINESTRING`, `GEOMETRYCOLLECTION` are rejected.
- **Validity**: each geometry should pass `sf::st_is_valid()`. Run `sf::st_make_valid()` on raw shapes before saving.
- **CRS**: the package does not enforce a specific CRS, but downstream raster operations and area computations assume metres. A common pattern is to re-project each layer to the country's UTM zone for `area`, then transform back to a display CRS (often EPSG:4326) before saving. Pick one CRS for the whole list and stick with it.
- **Area**: compute `area` in km² *after* projecting to UTM. `sf::st_area()` returns m² with a `units` attribute; convert with `units::set_units(., "km^2")` and drop the units attribute before saving.

### 1.6 Save format

Save the assembled list as compressed RDS:

```r
saveRDS(my_shp, "app-data/admin_bounds.rds", compress = "gz")
```

A compressed Ukraine-sized RDS is typically a few MB.

### 1.7 What `validate_geometries()` catches

`validate_geometries(my_shp)` runs the following checks per layer:

- Layer name parses to `admin<digit>_<name>`.
- Element is an `sf` object.
- Has a column called `geometry` of class `sfc`.
- All geometries are `POLYGON` or `MULTIPOLYGON`.
- `admin<N>Pcod` and `admin<N>Name` columns exist.
- `admin<N>Pcod` is unique (no duplicate IDs).
- `admin<N>Name` is unique (no duplicate names).
- No `NA`s in `admin<N>Pcod` or `admin<N>Name`.
- All parent-level P-code columns exist.
- All parent P-code values appear in their parent layer (cascade rule, §1.4).

It also verifies, across the full list, that `get_mt(my_shp)` produces a row for every polygon at the most-disaggregated level — i.e. that the levels form a strict hierarchy with no gaps.

### 1.8 What `validate_geometries()` does *not* check

- CRS consistency between layers.
- Whether `area` is in km² versus m² (it cannot tell).
- Whether geometries are topologically valid (use `sf::st_is_valid()` separately).
- Whether the polygons cover the country (gaps are not detected).

---

## 2. Metadata Excel template — `mtdt.xlsx`

The metadata template is a single `.xlsx` file with one sheet per admin level, plus three book-keeping sheets. `fct_template_reader()` parses it into the list shape that `launch_pti()` consumes.

### 2.1 Sheet inventory

| Sheet name        | Required?   | Purpose                                                                                                                                            |
| ----------------- | ----------- | -------------------------------------------------------------------------------------------------------------------------------------------------- |
| `general`         | yes         | One-row tibble with country-level metadata (`country` column).                                                                                      |
| `metadata`        | yes         | One row per indicator. Fourteen columns describing each indicator (see §2.3).                                                                       |
| `admin<N>_<Name>` | one or more | Wide-format indicator values. One column per indicator at that admin level. Sheet names must match the names in your `admin_bounds.rds` exactly.    |
| `weights_table`   | optional    | Pre-defined weight schemes. Each scheme is a `ws<digit>..` column group. See §2.5.                                                                  |
| `point_data`      | optional    | Reserved for future point-overlay support. Currently ignored by the pipeline; safe to omit.                                                         |

The bundled `ukr_mtdt_full` is the canonical example:

```{r}
data(ukr_mtdt_full)
names(ukr_mtdt_full)
```

The populated sheets here are `general`, `admin1_Oblast`, `admin2_Rayon`, `admin4_Hexagon`, and `metadata` — note that the admin-level sheet names match the slot names in `ukr_shp` exactly. This is mandatory.

### 2.2 The `general` sheet

A one-row tibble. The pipeline currently uses one column:

| Column    | Type      | Purpose                                          |
| --------- | --------- | ------------------------------------------------ |
| `country` | character | Display name of the country (e.g. `"Ukraine"`).  |

Other columns are tolerated but ignored.

```{r}
ukr_mtdt_full$general
```

### 2.3 The `metadata` sheet

This is the **indicator catalogue**. One row per indicator. Fourteen columns. Every column listed below is required (write a default if you have nothing meaningful for it).

| Column                  | Type        | Required    | Purpose                                                                                                                                            |
| ----------------------- | ----------- | ----------- | -------------------------------------------------------------------------------------------------------------------------------------------------- |
| `var_code`              | character   | yes         | The indicator's column name as it appears in the admin sheets. Must match exactly. **No `:`.**                                                     |
| `var_name`              | character   | yes         | Display name shown in the dashboard UI. **No `:`.**                                                                                                |
| `var_description`       | character   | recommended | Long-form description: how the indicator was constructed, sources used, units, caveats.                                                            |
| `var_order`             | integer     | yes         | Sort order within a pillar. Indicators are listed `pillar_group` then `var_order` ascending.                                                       |
| `var_units`             | character   | optional    | Unit of measurement (e.g. `"km"`, `"%"`, `"USD"`, `"persons/km²"`). Display-only.                                                                  |
| `spatial_level`         | character   | yes         | The most disaggregated admin level where this indicator has data. Value must equal an `admin<N>_<Name>` sheet name in this same file.              |
| `pillar_group`          | numeric     | yes         | Numeric code grouping indicators into pillars. All indicators with the same `pillar_group` form one pillar.                                         |
| `pillar_name`           | character   | yes         | Human-readable pillar name (e.g. `"Human capital"`). All rows sharing a `pillar_group` must share `pillar_name`. **No `:`.**                       |
| `pillar_description`    | character   | optional    | Long-form pillar description. Display-only.                                                                                                        |
| `fltr_exclude_pti`      | logical     | yes         | `TRUE` excludes this indicator from the PTI calculation. `FALSE` (default) includes it.                                                            |
| `fltr_exclude_explorer` | logical     | yes         | `TRUE` excludes this indicator from the Data Explorer tab. `FALSE` (default) includes it.                                                          |
| `fltr_overlay_pti`      | logical     | yes         | Reserved overlay flag. Set `FALSE` unless you know you need it.                                                                                    |
| `fltr_overlay_explorer` | logical     | yes         | Reserved overlay flag. Set `FALSE`.                                                                                                                |
| `legend_revert_colours` | logical     | yes         | `TRUE` reverses the colour scale for this indicator's legend (use when high values are *bad*).                                                     |

**Rules of thumb:**

- All `fltr_*` columns and `legend_revert_colours` must be **booleans**, not strings. If Excel renders them as `"TRUE"` / `"FALSE"` strings, `validate_read_metadata()` will complain.
- `var_code` values must be valid R column names (no spaces, no leading digits, no special characters except `_`).
- The `:` character breaks app compilation in `pillar_name` and `var_name` because it's used internally as a separator.

The bundled metadata sheet illustrates the layout:

```{r}
ukr_mtdt_full$metadata[, c(
  "var_code", "var_name", "spatial_level",
  "pillar_group", "pillar_name", "fltr_exclude_pti"
)]
```

### 2.4 Per-admin sheets

One sheet per admin level you have indicator data for. Sheet name must be **identical** to the corresponding slot name in `admin_bounds.rds` (e.g. `admin2_Rayon`). Wide format: rows are polygons, columns are indicators.

Required columns:

| Column                | Type      | Purpose                                                                                                              |
| --------------------- | --------- | -------------------------------------------------------------------------------------------------------------------- |
| `admin<N>Pcod`        | character | Foreign key. Must match `admin<N>Pcod` in the corresponding `sf` layer 1-to-1.                                        |
| `admin<N>Name`        | character | Mirror of the geometry's name column.                                                                                 |
| Parent `admin<k>Pcod` | character | All `k < N` ancestor P-codes (cascade rule, §1.4).                                                                    |
| `year`                | integer   | Vintage of the values in this row. May vary across rows; pipeline aggregates by P-code rather than year.              |
| `area`                | numeric   | Area in km², usually copied from the geometry layer.                                                                  |

Then **one column per indicator measured at this level**. Column name must equal the indicator's `var_code` in the `metadata` sheet. Type should be `numeric` (`integer` is OK).

`fct_template_reader()` drops indicator columns that are not declared in the `metadata` sheet, and drops admin sheets that have no remaining indicator columns. So the system tolerates "extra" columns in your raw data — they are silently filtered out.

```{r}
str(ukr_mtdt_full$admin1_Oblast)
```

### 2.5 The optional `weights_table` sheet

Lets you ship pre-defined weight schemes alongside the dashboard. Users can pick from these in the UI instead of building weights manually.

Layout: one row per indicator (`var_code`), then column groups named `ws<digit>..<column>` — one group per weight scheme. Each scheme has at least:

| Column             | Type      | Purpose                                                                                                          |
| ------------------ | --------- | ---------------------------------------------------------------------------------------------------------------- |
| `ws<digit>..name`  | character | Display name of the weight scheme (e.g. `"Equal weights"`, `"Poverty-heavy"`).                                   |
| `ws<digit>..value` | numeric   | The weight assigned to this `var_code` under this scheme. Sum within a scheme is normalised internally.          |

Multiple schemes use successive `ws1..`, `ws2..`, `ws3..` prefixes. `fct_convert_weight_to_clean()` (internal) reshapes this into the per-scheme list the calculator consumes.

Omit the sheet entirely if you want users to build all weights from the UI.

### 2.6 What `validate_metadata()` catches

`validate_metadata(shp_path, mtdt_path)` runs the file through `validate_read_shp()`, then `validate_read_metadata()`, then executes the full calculation pipeline under all-equal weights and confirms the number of scored pillars equals the number of indicators. Specifically:

- The shapes file is readable as a non-empty named list (`validate_read_shp()`).
- The metadata file is readable as a non-empty `metadata` sheet (`validate_read_metadata()`).
- All `fltr_*` columns are read as logical, not string.
- The pipeline runs without error and produces the expected count of scored pillars.

### 2.7 What `validate_metadata()` does *not* check (yet)

These are tracked under issue [#7](https://github.com/worldbank/devPTIpack/issues/7) and will be added incrementally:

- ID-mismatch reports between `admin<N>Pcod` in your sheets and the `sf` layer.
- Per-indicator data-type and NA-percentage summaries.
- CSV-format inputs (currently `.xlsx` only).
- Auto-generated metadata from a bare CSV.

---

## 3. Common pitfalls

A non-exhaustive list of things that have bitten developers before. Most are caught by the validators; a few are not.

1. **`fltr_*` stored as strings.** Excel's autoformat sometimes saves `TRUE`/`FALSE` as text. Set the cell type to *Boolean* (or use `writexl::write_xlsx()` from R with logical columns) so `fct_template_reader()` reads them correctly.
2. **Indicator column not in `metadata` sheet.** If the indicator column exists in `admin1_Oblast` but is not declared in the `metadata` sheet, it is silently dropped at template load. App users will not see it.
3. **`var_code` mismatch.** A typo between the indicator column name in the admin sheet and `var_code` in the metadata sheet means the indicator is dropped without an error.
4. **`spatial_level` not matching a sheet.** If the metadata says `spatial_level = "admin3_District"` but no such sheet exists, the indicator is dropped.
5. **`admin0_Country` missing.** The pipeline requires the country polygon even if the UI only displays admin1+.
6. **Lower-case admin keys.** `admin1_oblast` ≠ `admin1_Oblast`. Names are case-sensitive across `admin_bounds.rds` slots, `metadata$spatial_level` values, and admin sheet names.
7. **`:` in pillar or variable names.** The package uses `:` internally as a separator; `:` in display strings will fragment them.
8. **Pillar metadata inconsistent.** Two rows with the same `pillar_group` must share the same `pillar_name` and `pillar_description`, or rendering becomes ambiguous.
9. **Area in m² not km².** `validate_geometries()` cannot detect this; the legend just shows surprisingly large numbers. Always convert with `units::set_units(., "km^2")` before saving.
10. **CRS mismatch between layers.** The package does not enforce a single CRS. Inconsistent CRSs can produce wrong centroids and wrong `area`. Re-project once at the top of your data-prep pipeline and reuse.
11. **Orphan child P-codes.** `admin2_District$admin1Pcod` values must all appear in `admin1_Region$admin1Pcod`. `validate_geometries()` warns on orphans.
12. **Non-numeric indicator columns.** Indicator columns must be numeric. Strings, factors, dates are rejected by the pipeline.

---

## 4. Worked end-to-end example

This example assembles a metadata Excel file from the bundled data, then validates it. Runnable in any session that has `devPTIpack` installed.

```{r eval=FALSE}
library(devPTIpack)
library(writexl)

data(ukr_shp)
data(ukr_mtdt_full)

# 1. Save the bundled boundary list as the kind of .rds a user would ship.
shp_path <- tempfile(fileext = ".rds")
saveRDS(ukr_shp, shp_path)

# 2. Build a metadata template by lifting the bundled metadata + admin sheets
#    into a list of data frames in the shape writexl expects.
sheets <- list(
  general       = ukr_mtdt_full$general,
  metadata      = ukr_mtdt_full$metadata,
  admin1_Oblast = ukr_mtdt_full$admin1_Oblast,
  admin2_Rayon  = ukr_mtdt_full$admin2_Rayon
)

mtdt_path <- tempfile(fileext = ".xlsx")
writexl::write_xlsx(sheets, mtdt_path)

# 3. Read it back through the package's template reader to confirm the
#    round-trip works.
parsed <- fct_template_reader(mtdt_path)
names(parsed)

# 4. Validate the geometries (in-memory).
validate_geometries(ukr_shp)

# 5. Validate the on-disk pair (shapes + metadata) end-to-end.
validate_metadata(shp_path = shp_path, mtdt_path = mtdt_path)
```

If both validators report `pass`, you can launch the app:

```r
launch_pti(
  shp_dta  = readRDS(shp_path),
  inp_dta  = fct_template_reader(mtdt_path),
  app_name = "Worked example"
)
```

---

## See also

- [Build a PTI](articles/build-pti.html) — the task-oriented walkthrough this page is referenced from.
- [PTI Methodology](articles/methodology.html) — the conceptual framing (what PTI is, why it exists, the math).
- [`?validate_geometries`](reference/validate_geometries.html), [`?validate_metadata`](reference/validate_metadata.html) — the authoritative validator references.
- [`?ukr_shp`](reference/ukr_shp.html), [`?ukr_mtdt_full`](reference/ukr_mtdt_full.html) — the bundled sample inputs that all examples on this page use.

External worked-example repositories (private — request access from the listed maintainers):

- [`zamPTIdata`](https://github.com/wbPTI/zamPTIdata) — Zambia data-prep pipeline. Contact: [Taka Masaki](mailto:tmasaki@worldbank.org).
- `kgzPTIdata` — Kyrgyz Republic data-prep pipeline. Contact the package maintainer.
