---
title: "Step 1 -- Prepare shapefiles"
format: html
---

<!-- For detailed instructions, see:
     https://worldbank.github.io/devPTIpack/articles/build-pti-1-shapefiles.html -->

This step loads the Rwanda boundary GeoJSONs from `sample-data/`,
attaches the package's `admin<N>Pcod` / `admin<N>Name` / `area`
columns, runs `validate_geometries()`, and saves the result to
`app-data/shapes.rds` for use by the Shiny app and downstream steps.

Replace the GeoJSON paths and the `admin<N>_<HumanName>` slot names
with your own country's data when adapting this template.

## Load raw shapefiles

```{r}
#| message: false
library(sf)
library(dplyr)
library(devPTIpack)

adm0_raw <- read_sf("sample-data/rwa_adm0.geojson")
adm1_raw <- read_sf("sample-data/rwa_adm1.geojson")
adm2_raw <- read_sf("sample-data/rwa_adm2.geojson")
```

## Attach package-convention columns

The package expects every layer to carry `admin<N>Pcod` (unique polygon
code), `admin<N>Name` (human-readable name) and `area` columns, plus
parent-level P-codes on sub-admin layers (so admin2 carries an
`admin1Pcod` column that joins back to admin1). geoBoundaries gives us
`shapeID` / `shapeName` per layer; we rename those and derive the
admin1 parent for each admin2 polygon via a centroid spatial join.

```{r}
adm0 <- adm0_raw |>
  mutate(
    admin0Pcod = "RWA",
    admin0Name = shapeName,
    area       = as.numeric(st_area(geometry))
  ) |>
  select(admin0Pcod, admin0Name, area, geometry)

adm1 <- adm1_raw |>
  mutate(
    admin0Pcod = "RWA",
    admin1Pcod = shapeID,
    admin1Name = shapeName,
    area       = as.numeric(st_area(geometry))
  ) |>
  select(admin0Pcod, admin1Pcod, admin1Name, area, geometry)

# Spatial join admin2 centroids -> admin1 to derive each district's
# parent province code.
adm2_centroids <- suppressWarnings(st_centroid(adm2_raw))
parent_lookup  <- suppressWarnings(
  st_join(adm2_centroids, adm1_raw, join = st_within, suffix = c("", "_p"))
) |>
  st_drop_geometry() |>
  transmute(child_id = shapeID, admin1Pcod = shapeID_p)

adm2 <- adm2_raw |>
  left_join(parent_lookup, by = c("shapeID" = "child_id")) |>
  mutate(
    admin0Pcod = "RWA",
    admin2Pcod = shapeID,
    admin2Name = shapeName,
    area       = as.numeric(st_area(geometry))
  ) |>
  select(admin0Pcod, admin1Pcod, admin2Pcod, admin2Name, area, geometry)
```

## Assemble the shapes list

Slot names follow `admin<N>_<HumanName>` -- the level prefix is what
downstream code parses to build the admin hierarchy. The human-name
suffix is free-form (e.g. `Country`, `Province`, `District`).

```{r}
my_shp <- list(
  admin0_Country  = adm0,
  admin1_Province = adm1,
  admin2_District = adm2
)
```

## Validate

```{r}
validate_geometries(my_shp)

# TODO #80 -- visual inspection via standalone Shiny app.
# app_validate_shp() lands in issue #80; uncomment when that PR merges.
# app_validate_shp(my_shp)
```

## Save to app-data/

```{r}
dir.create("app-data", showWarnings = FALSE)
saveRDS(my_shp, "app-data/shapes.rds")
```
