---
title: "Step 3 -- Prepare metadata Excel"
format: html
---

<!-- For detailed instructions, see:
     https://worldbank.github.io/devPTIpack/articles/build-pti-3-metadata.html -->

This step reads the synthetic Rwanda metadata workbook from
`sample-data/`, validates it against the shapefile saved in Step 1,
and copies the workbook to `app-data/metadata-user.xlsx` so it can be
picked up by Step 5's compile pass.

The Rwanda template ships two sample workbooks:

- `sample-metadata-adm1.xlsx` -- 5 provinces x 3 indicators (simple
  one-level case).
- `sample-metadata-adm1-adm2.xlsx` -- same 3 indicators at both Adm1
  and Adm2 (multi-level case).

This step uses the multi-level workbook by default; switch to the
Adm1-only workbook by changing the `mtdt_path` variable below.

## Read the metadata template

```{r}
#| message: false
library(devPTIpack)

mtdt_path <- "sample-data/sample-metadata-adm1-adm2.xlsx"
inp_dta   <- fct_template_reader(mtdt_path)
names(inp_dta)
```

## Validate against your shapefile

```{r}
validate_metadata(
  shp_path  = "app-data/shapes.rds",
  mtdt_path = mtdt_path
)

# TODO #81 -- visual inspection via standalone Shiny app.
# app_validate_metadata() lands in issue #81; it pairs the validator
# output with the bundled mod_dta_explorer2 modules so you can scan
# every indicator's values on the map even when validation fails.
# Uncomment when that PR merges.
# shp_dta <- readRDS("app-data/shapes.rds")
# app_validate_metadata(shp_dta = shp_dta, inp_dta = inp_dta)
```

## Stage the workbook for Step 5

`compile_pti_data()` (Step 5) merges every intermediate metadata file
under `app-data/`. Copy the validated user workbook to its canonical
location now:

```{r}
dir.create("app-data", showWarnings = FALSE)
file.copy(
  from      = mtdt_path,
  to        = "app-data/metadata-user.xlsx",
  overwrite = TRUE
)
```
