---
title: "Step 4 -- HEX data"
format: html
---

<!-- For detailed instructions, see:
     https://worldbank.github.io/devPTIpack/articles/build-pti-4-hex.html -->

::: {.callout-warning}
**Feature not ready.** This step depends on the hex data access pipeline
(arch-11), which is under active development. The functions referenced
below do not yet exist. This file ships as a working skeleton — uncomment
and adapt when the pipeline lands.
:::

## Configuration

The `HEX_RESOLUTION` and `INCLUDE_HEX_IN_APP` switches are set in
`00-master.R`. `HEX_RESOLUTION` controls what H3 resolution was used
to build `admin9_Hexagon` in Step 1 — `fetch_hex_data()` reads it
automatically from the hex IDs. `INCLUDE_HEX_IN_APP` controls whether
hex-level polygons are included in the deployed app.

```{r}
#| eval: false
# Inherited from 00-master.R:
# HEX_RESOLUTION    <- 6L     # 5 = ~252 km², 6 = ~36 km², 7 = ~5 km²
# INCLUDE_HEX_IN_APP <- FALSE
```

## Load shapes and hex layer

```{r}
#| eval: false
library(devPTIpack)

my_shp    <- readRDS("app-data/shapes.rds")
hex_layer <- my_shp$admin9_Hexagon
hex_ids   <- hex_layer$admin9Pcod
```

## A. Discover available variables

```{r}
#| eval: false
# Browse the full registry catalogue
list_hex_vars()

# Check live available years for a specific variable
get_available_years(vars = NULL, "nightlights")
```

## B. Configure variable selection

```{r}
#| eval: false
# Population is always fetched automatically (needed for weighted
# aggregation). Do not list it here — only list indicator variables.
vars <- use_hex_vars(
  "nightlights",
  "poverty_rate",
  years = c(2015, 2022)
)
```

## C. Fetch hex data

```{r}
#| eval: false
hex_data <- fetch_hex_data(hex_ids = hex_ids, vars = vars)
```

## C′. Optional: merge your own local parquet

```{r}
#| eval: false
# Uncomment and adapt if you have project-specific indicators.
# Local parquet must be long format, no population column.
#
# my_data <- arrow::open_dataset("data-raw/my_indicators.parquet") |>
#   dplyr::filter(my_hex_col %in% hex_ids) |>
#   dplyr::collect() |>
#   dplyr::rename(hex_id = my_hex_col) |>
#   tidyr::pivot_wider(
#     names_from  = year,
#     values_from = my_variable,
#     names_prefix = "my_var_"
#   )
#
# hex_data <- dplyr::full_join(hex_data, my_data, by = "hex_id")
```

## D. Aggregate to admin polygons

```{r}
#| eval: false
aggregated <- aggregate_hex_to_shapes(
  hex_data  = hex_data,
  hex_layer = hex_layer,
  shp_dta   = my_shp,
  strategy  = list(
    # population is always aggregated internally as weight="none", fun="sum".
    # Do not set it here.
    .default    = c(weight = "pop",  fun = "mean"),
    nightlights = c(weight = "none", fun = "mean")
  )
)
```

## E. Build metadata Excel

```{r}
#| eval: false
build_hex_metadata(
  aggregated       = aggregated,
  shp_dta          = my_shp,
  # indicator_config needed only for non-registry variables:
  # indicator_config = tibble::tibble(
  #   canonical_name = "my_var",
  #   var_name       = "My Variable ({year})",
  #   pillar_group   = 4,
  #   pillar_name    = "Custom"
  # ),
  country_name       = "Rwanda",
  output_path        = "app-data/metadata-hex.xlsx",
  include_hex        = INCLUDE_HEX_IN_APP,
  # include_population = TRUE  # uncomment to add population as a visible
  #                            # map indicator (excluded by default to avoid
  #                            # collision with Step 3 population variables)
)
```

## Ordering note

This step is **order-flexible** -- it can be run before or after
Step 3. The two outputs (`metadata-user.xlsx` and `metadata-hex.xlsx`)
are independent intermediate files merged together in Step 5.
