---
title: "Step 5 -- Compile and finalise"
format: html
---

<!-- For detailed instructions, see:
     https://worldbank.github.io/devPTIpack/articles/build-pti-5-compile.html -->

::: {.callout-warning}
**Function not yet implemented.** `compile_pti_data()` is delivered by
issue #83 (branch `feat/compile-pti-data`). Until that PR merges, this
step is a stub: the call below is commented out, and `00-master.R`
does **not** render this file by default.
:::

## What this step will do

Once `compile_pti_data()` lands, this step will:

1. Read every `metadata-*.xlsx` intermediate file produced by Steps 3
   and 4 from `app-data/`.
2. Merge them into a single canonical `app-data/metadata.xlsx`.
3. Re-run `validate_geometries()` and `validate_metadata()` on the
   combined inputs.
4. Render `app-data/pti-metadata.pdf` -- the printable indicator
   reference, mapping each indicator at its most-disaggregated admin
   level (or at the explicit `spatial_level` you set in metadata).
5. Export `app-data/shapefiles.zip` -- a GeoJSON archive (one file per
   admin level) for download from the deployed app.
6. Print a verbose CLI summary (admin levels, polygon counts,
   indicator and pillar counts, files produced).

## Run

```{r}
#| eval: false
# TODO #83 -- compile_pti_data() not yet implemented.
# Uncomment after issue #83 lands.
#
# compile_pti_data(
#   shp_path       = "app-data/shapes.rds",
#   metadata_paths = c(
#     "app-data/metadata-user.xlsx"
#     # "app-data/metadata-hex.xlsx"  # add when Step 4 produces output
#   ),
#   output_dir     = "app-data"
# )
```

## Pre-launch checklist

After `compile_pti_data()` completes you should see all of:

- `app-data/shapes.rds`
- `app-data/metadata.xlsx`
- `app-data/pti-metadata.pdf`
- `app-data/shapefiles.zip`

Edit `landing-page.md` (in the project root) to describe your app, then
launch locally with:

```{r}
#| eval: false
# devPTIpack::launch_pti(
#   shp_dta = readRDS("app-data/shapes.rds"),
#   inp_dta = devPTIpack::fct_template_reader("app-data/metadata.xlsx")
# )
```
