---
title: "Step 5 -- Compile and finalise"
format: html
---

<!-- For detailed instructions, see:
     https://worldbank.github.io/devPTIpack/articles/build-pti-5-compile.html -->

This step takes the intermediate metadata files produced by Steps 3
(and optionally Step 4) and `compile_pti_data()` produces the three
deployment-ready artefacts under `app-data/`:

- `app-data/metadata.xlsx` -- canonical merged metadata workbook.
- `app-data/shapefiles.zip` -- GeoJSON archive (one file per admin
  level) for download from the deployed app.
- `app-data/pti-metadata.pdf` -- printable indicator reference,
  mapping each indicator at its most-disaggregated admin level (or
  at the explicit `spatial_level` you set in metadata).

`compile_pti_data()` also re-runs `validate_geometries()` and
`validate_metadata()` on the combined inputs and prints a verbose
CLI summary (layers, polygon counts, indicator + pillar counts,
files produced). On validation failure it errors by default; pass
`error_on_fail = FALSE` to inspect the structured result instead.

## Run

```{r}
#| eval: false
library(devPTIpack)

compile_pti_data(
  shp_path       = "app-data/shapes.rds",
  metadata_paths = c(
    "app-data/metadata-user.xlsx"
    # "app-data/metadata-hex.xlsx"  # uncomment when Step 4 produces output
  ),
  output_dir     = "app-data"
)
```

## Pre-launch checklist

After `compile_pti_data()` completes you should see all of:

- `app-data/shapes.rds`
- `app-data/metadata.xlsx`
- `app-data/pti-metadata.pdf` (skipped with a warning if LaTeX is unavailable)
- `app-data/shapefiles.zip`

Edit `landing-page.md` (in the project root) to describe your app, then
launch locally:

```{r}
#| eval: false
launch_pti(
  shp_dta  = readRDS("app-data/shapes.rds"),
  inp_dta  = fct_template_reader("app-data/metadata.xlsx"),
  app_name = "My country PTI"
)
```
