---
title: "Build a PTI"
description: >
  A step-by-step walkthrough of taking raw country data through to a deployed
  Project Targeting Index (PTI) Shiny dashboard with `devPTIpack`.
---

```{r setup_lib, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

This is the recipe. If you've never built a PTI app before, follow it top to bottom — at the end you will have a Shiny dashboard running locally that you can deploy to a server.

This page complements [PTI Methodology](methodology.html) (the *why*) and [Data preparation reference](dataprep.html) (the canonical *what*). The walkthrough below points into the reference whenever a question of input shape comes up; you don't need to read the reference cover-to-cover before starting.

**Time estimate.** With clean shapes and a tidy indicator dataset already in hand, a developer who has done this before launches an app in roughly an hour. Most of the time on a new country goes into preparing the data, not configuring the app.

---

## 0. What you'll build

By the end of this guide you will have a project directory shaped roughly like this:

```
myPTI/
├── app.R                           # the launcher
├── app-data/
│   ├── admin_bounds.rds            # boundary shapes (this is what you build)
│   ├── mtdt.xlsx                   # metadata template (this is what you build)
│   ├── metadata.pdf                # auto-rendered indicator catalogue (optional)
│   └── shapefiles.zip              # raw shapes for download (optional)
├── R/
│   └── _disable_autoload.R         # golem boilerplate
├── landing-page.md                 # info-modal content (optional)
└── myPTI.Rproj
```

Running `app.R` launches a multi-tab Shiny dashboard with an interactive map of every administrative level you ship, weighted PTI scores per polygon, a Data Explorer tab, and a Compare tab.

---

## 1. Prerequisites

You'll need:

- R `>= 4.1.0`.
- The `devPTIpack` package (and its dependencies — `sf`, `shiny`, `golem`, `leaflet`, etc., installed transitively).
- Administrative-boundary shapefiles for every level you want to display (typically admin0–admin2 or admin0–admin3).
- One or more indicator datasets, each indicator measured at one specific admin level.

Install the package from GitHub:

```r
remotes::install_github("worldbank/devPTIpack")
```

Smoke-test the install with the bundled Ukraine sample data — if this command opens a working dashboard, your environment is ready:

```r
library(devPTIpack)
launch_pti_onepage(
  shp_dta  = ukr_shp,
  inp_dta  = ukr_mtdt_full,
  app_name = "Smoke test"
)
```

Once that works, you can move on to building your own.

---

## 2. Scaffold the project

`create_new_pti()` drops a minimal app skeleton into a new directory:

```r
library(devPTIpack)
create_new_pti("path/to/myPTI")
```

This creates the directory, copies in `app.R`, an empty `R/` folder, a `landing-page.md` placeholder, and (if you're in RStudio) opens the new project automatically. The skeleton's `app.R` is wired up to the bundled Ukraine data initially — change those references to your own paths once you have your inputs.

You will also want an `app-data/` sub-directory inside the project. Create it manually — the scaffolder doesn't create it because the contents are entirely yours to provide:

```r
dir.create(file.path("path/to/myPTI", "app-data"))
```

The scaffolded `app.R` looks like this:

```r
library(shiny)
library(devPTIpack)

options("golem.app.prod" = TRUE)
launch_pti(
  shp_dta         = devPTIpack::ukr_shp,
  inp_dta         = devPTIpack::ukr_mtdt_full,
  app_name        = "{{COUNTRY NAME}}",
  show_waiter     = TRUE,
  show_adm_levels = NULL,
  shapes_path     = ".",
  mtdtpdf_path    = "."
)
```

You'll edit `shp_dta`, `inp_dta`, `app_name`, and the path arguments in §6 once your data is ready.

---

## 3. Prepare boundary shapes

The package expects a **named list of `sf` tibbles**, one per administrative level, saved as a single `.rds` file. The full shape specification (column requirements, parent-child cascade rule, geometry constraints) is in [Data preparation reference §1](dataprep.html#boundary-shapes-admin_bounds.rds). Read it before assembling your list — most validator failures originate in that section's invariants.

The minimal recipe, assuming you have raw shapefiles per admin level:

```r
library(sf)
library(dplyr)
library(units)

# Read every level. For your country, replace these paths.
admin1 <- read_sf("data-raw/admin1.shp")
admin2 <- read_sf("data-raw/admin2.shp")

# Re-project to a metric CRS for area; here we use UTM 36N (Ukraine).
# Pick a projection appropriate for your country.
to_metric <- function(x) st_transform(x, 32636)

admin1 <- admin1 |> to_metric() |> st_make_valid()
admin2 <- admin2 |> to_metric() |> st_make_valid()

# Compute area in km².
admin1$area <- as.numeric(set_units(st_area(admin1), "km^2"))
admin2$area <- as.numeric(set_units(st_area(admin2), "km^2"))

# Derive admin0 from admin1 by dissolving on the country code.
admin0 <- admin1 |>
  mutate(admin0Pcod = "UA", admin0Name = "Ukraine") |>
  group_by(admin0Pcod, admin0Name) |>
  summarise(area = sum(area), .groups = "drop") |>
  st_make_valid()

# Each layer must contain its own Pcod/Name plus all parent Pcods.
# admin1 must carry admin0Pcod; admin2 must carry admin0Pcod and admin1Pcod.
# (Your raw shapefiles usually do — verify column names match the package
# convention exactly; rename if needed.)

# Assemble the named list. Names follow `admin<N>_<HumanName>` exactly.
my_shp <- list(
  admin0_Country = admin0,
  admin1_Oblast  = admin1,
  admin2_Rayon   = admin2
)

# Save it.
saveRDS(my_shp, "app-data/admin_bounds.rds", compress = "gz")
```

For the column-by-column rules (which columns each layer needs, what the cascade rule says), see [Data preparation reference §1.3–§1.4](dataprep.html#required-columns-per-layer).

---

## 4. Prepare the metadata template

The metadata template is a single `.xlsx` file with one sheet per admin level, plus a `general` sheet, plus a `metadata` sheet describing each indicator. The full sheet-by-sheet and column-by-column reference is in [Data preparation reference §2](dataprep.html#metadata-excel-template-mtdt.xlsx).

The high-level pattern: build each sheet as a data frame in R, collect them into a named list, write the list to `.xlsx` with `writexl::write_xlsx()`. This is the same pattern documented in the [`zamPTIdata`](https://github.com/wbPTI/zamPTIdata) and [`kgzPTIdata`](https://github.com/wbPTI/kgzPTIdata) example repositories (both private — request access from [Eduard Bukin](https://github.com/EBukin)).

A skeletal version, joining indicator data onto the boundary list from §3:

```r
library(dplyr)
library(sf)
library(writexl)

# Suppose you have indicator tibbles already aggregated to the right level —
# one indicator per column, plus the keys.
poverty_admin2 <- ...        # cols: admin2Pcod, admin2Name, year, poverty_rate
literacy_admin1 <- ...       # cols: admin1Pcod, admin1Name, year, literacy_rate

# 1. Build the per-admin sheets by joining indicators onto the boundary keys.
admin1_sheet <- my_shp$admin1_Oblast |>
  st_drop_geometry() |>
  left_join(literacy_admin1, by = c("admin1Pcod", "admin1Name"))

admin2_sheet <- my_shp$admin2_Rayon |>
  st_drop_geometry() |>
  left_join(poverty_admin2, by = c("admin2Pcod", "admin2Name"))

# 2. Build the metadata sheet — one row per indicator. See dataprep §2.3 for
#    the full column reference.
metadata <- tibble::tribble(
  ~var_code,        ~var_name,             ~var_description,                       ~var_order, ~var_units, ~spatial_level,    ~pillar_group, ~pillar_name,         ~pillar_description, ~fltr_exclude_pti, ~fltr_exclude_explorer, ~fltr_overlay_pti, ~fltr_overlay_explorer, ~legend_revert_colours,
  "literacy_rate",  "Literacy rate",       "Adult literacy rate, latest census.",  1L,         "%",        "admin1_Oblast",    1,             "Human capital",      NA,                  FALSE,             FALSE,                  FALSE,             FALSE,                  FALSE,
  "poverty_rate",   "Poverty headcount",   "Share of population below the line.",  2L,         "%",        "admin2_Rayon",     1,             "Human capital",      NA,                  FALSE,             FALSE,                  FALSE,             FALSE,                  TRUE
)

# 3. The general sheet.
general <- tibble::tibble(country = "Ukraine")

# 4. Collect into a named list and write.
sheets <- list(
  general       = general,
  metadata      = metadata,
  admin1_Oblast = admin1_sheet,
  admin2_Rayon  = admin2_sheet
)

writexl::write_xlsx(sheets, "app-data/mtdt.xlsx")
```

A few rules of thumb (the full list is in [dataprep §3](dataprep.html#common-pitfalls)):

- The `var_code` in `metadata` must equal an indicator column name in at least one admin sheet, exactly.
- `spatial_level` must equal one of the admin sheet names in this same file (not the geometry list, though they should be identical).
- All `fltr_*` columns and `legend_revert_colours` are *booleans*, not strings.
- Don't put `:` in `var_name` or `pillar_name`.

---

## 5. Validate the inputs

Before launching the app, run the two exported validators. They emit pass/warn/fail messages and (since the validator UX pass) return an invisible structured result you can inspect.

```r
library(devPTIpack)

my_shp <- readRDS("app-data/admin_bounds.rds")

validate_geometries(my_shp)
validate_metadata(
  shp_path  = "app-data/admin_bounds.rds",
  mtdt_path = "app-data/mtdt.xlsx"
)
```

`validate_geometries()` walks every layer and reports per-layer findings (geometry types, P-code uniqueness, name uniqueness, parent-child cascade). `validate_metadata()` runs both file-readability checks and a dry-run of the calculation pipeline under all-equal weights.

A clean run prints a list of pass alerts. Warnings are non-blocking (the app will still launch but something is suspect). Fails are blocking (the pipeline will not run). Each issue includes the offending layer / column / value so you can fix it surgically.

If you're scripting validation programmatically (for example inside CI), capture the structured return:

```r
res <- validate_geometries(my_shp, error_on_fail = FALSE)
res$status     # "pass", "warn", or "fail"
res$issues     # list of issue records — level, check, message, details
```

The `error_on_fail = FALSE` argument lets the function return cleanly on a failure instead of throwing; default is `TRUE` so existing pipeline scripts that depended on the validator aborting still work.

For the full list of what the validators do and don't check, see [dataprep §1.7–§1.8](dataprep.html#what-validate_geometries-catches) and [dataprep §2.6–§2.7](dataprep.html#what-validate_metadata-catches).

---

## 6. Launch locally

Edit `app.R` to point at your inputs:

```r
library(shiny)
library(devPTIpack)

options("golem.app.prod" = TRUE)
launch_pti(
  shp_dta         = readRDS("app-data/admin_bounds.rds"),
  inp_dta         = fct_template_reader("app-data/mtdt.xlsx"),
  app_name        = "My Country PTI",
  show_waiter     = TRUE,
  show_adm_levels = c("admin1", "admin2"),
  shapes_path     = "app-data/shapefiles.zip",
  mtdtpdf_path    = "app-data/metadata.pdf"
)
```

Then click *Run App* in RStudio (or `shiny::runApp("path/to/myPTI")`). The app should open in a browser tab.

The most useful `launch_pti()` arguments:

| Argument          | Meaning                                                                                                                                                       |
| ----------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| `shp_dta`         | Required. The shapes list (or its on-disk RDS read with `readRDS()`).                                                                                          |
| `inp_dta`         | Required. The parsed metadata list (run your `mtdt.xlsx` through `fct_template_reader()`).                                                                     |
| `app_name`        | Display title in the navbar and downloads.                                                                                                                     |
| `ui_type`         | `"twocol"` (default — full multi-tab dashboard) or `"box"` (compact single-card embedding for putting PTI inside another Shiny app).                           |
| `show_adm_levels` | Vector like `c("admin1", "admin2")` controlling which levels users can compute PTI at. `NULL` (default) shows all levels with data.                            |
| `tabs`            | Subset of `c("info", "compare", "explorer")`. Drop `"info"` to skip the landing-page modal; drop `"compare"` to hide the side-by-side comparison view.         |
| `shapes_path`     | Path to a zip archive of raw shapefiles served as a download from the app. If `NULL`, the app generates one from `shp_dta` on the fly.                         |
| `mtdtpdf_path`    | Path to a pre-rendered PDF of the metadata catalogue, served as a download. If `NULL`, the app does not offer a PDF download.                                  |
| `show_waiter`     | `TRUE` to show a loading splash on cold start. Set `FALSE` when embedding the PTI as a tab inside a multi-tab dashboard.                                       |
| `pti_landing_page` | Path to a markdown file rendered as the landing-page modal. If omitted, the modal is skipped.                                                                  |

For one-page embedding (no navbar, no Compare/Explorer tabs), use `launch_pti_onepage()` — same `shp_dta` / `inp_dta` arguments, simpler UI.

---

## 7. Deploy

Three common targets. Pick whichever your team already runs.

**shinyapps.io** — easiest for a public demo.

```r
rsconnect::deployApp(appDir = "path/to/myPTI", appName = "myPTI")
```

Make sure `app.R`'s paths are *relative* (e.g. `"app-data/admin_bounds.rds"`, not `"~/projects/myPTI/app-data/..."`). Absolute paths break on the server.

**Posit Connect** — for internal-network deployments. Use the *Publish* blue-arrow button in RStudio with a Connect server configured.

**Self-hosted Shiny Server** — `scp` the project directory into `/srv/shiny-server/myPTI` (or wherever the server reads). The dependencies on the host need to match what's in `DESCRIPTION`.

For a real-world deployment recipe, see the [`zamPTI`](https://github.com/wbPTI/zamPTI) example app repo (private — request access from [Eduard Bukin](https://github.com/EBukin)).

---

## 8. Troubleshooting

| Symptom                                                                       | Likely cause                                                                                              | Fix                                                                                                          |
| ----------------------------------------------------------------------------- | --------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------------------------------------------------ |
| App launches but the indicator dropdown is empty                              | `var_code` in `metadata` doesn't match a column name in any `admin<N>_*` sheet                            | Verify column names letter-for-letter; remember case sensitivity                                              |
| `validate_read_metadata()` warns "filter columns are not read as logical"     | `fltr_*` cells stored as `"TRUE"` / `"FALSE"` strings                                                     | Re-write the `.xlsx` from R with `writexl::write_xlsx()` — the columns will then be true booleans            |
| `validate_geometries()` warns "values in `admin1Pcod` of `admin2_X` not in admin1_Y" | Orphan child P-codes — the cascade rule (dataprep §1.4) is broken                                          | Fix the upstream join, or filter the orphan rows out                                                          |
| Map renders but some polygons are blank (no colour)                           | The polygon has `NA` for every selected indicator                                                          | `get_vars_un_avbil()` lists unavailable admins. Either fill the data or accept the gap.                       |
| `launch_pti()` aborts with "indicator column type"                            | An indicator column is character / factor / Date instead of numeric                                       | Cast in the source pipeline; the package does not coerce silently                                             |
| Deployed app shows a 404 for the PDF download                                 | `mtdtpdf_path` is absolute or the file was not uploaded                                                    | Use a relative path; verify `app-data/metadata.pdf` exists in the deployed bundle                            |
| `validate_metadata()` complains about pillar count                            | Two indicators with the same `pillar_group` have different `pillar_name` strings                          | Make `pillar_name` consistent within each group                                                              |

If you hit something not on the list, run the validators with `error_on_fail = FALSE` and inspect `res$issues` — every check the package enforces is named in the `check` field.

---

## 9. Next steps

- [PTI Methodology](methodology.html) — the conceptual framing (z-scores, weights, monitoring).
- [Data preparation reference](dataprep.html) — the canonical input-shape spec referenced throughout this walkthrough.
- [Function reference](../reference/index.html) — every exported function, organised by category.
- [Past projects](past-projects.html) — what completed PTI deployments look like.

---

```{r, include=FALSE}
# Verify the bundled-data smoke test from §1 actually works (without
# launching the app — we only want to confirm the inputs parse).
library(devPTIpack)
data(ukr_shp)
data(ukr_mtdt_full)
stopifnot(is.list(ukr_shp), length(ukr_shp) > 0)
stopifnot(is.list(ukr_mtdt_full), "metadata" %in% names(ukr_mtdt_full))
```
