---
title: "Step 4 — HEX data"
---

Build the hex-derived indicator dataset and the companion metadata Excel
(`app-data/metadata-hex.xlsx`) that Step 5's `compile_pti_data()` merges
with your user metadata from Step 3.

The pipeline runs **four stages** driven by two `00-master.R` switches:

```r
HEX_RESOLUTION    <- 6L     # 5 ≈ 252 km², 6 ≈ 36 km², 7 ≈ 5 km²
INCLUDE_HEX_IN_APP <- FALSE
```

`HEX_RESOLUTION` controls the H3 resolution used when building
`admin9_Hexagon` in Step 1.  `fetch_hex_data()` reads the resolution
from your hex IDs automatically — you do not need to pass it explicitly.

`INCLUDE_HEX_IN_APP` controls whether the raw hex-level sheet is
written into `metadata-hex.xlsx` and shipped with the app.  Aggregated
indicator values at admin1, admin2, etc. are **always** included
regardless of this switch (see [§E](#e-build-metadata-excel) below).

This step is **order-flexible**: it can run before or after Step 3.
The outputs are independent intermediate files merged in Step 5.

## Prerequisites

Step 1 must be complete.  Your `shapes.rds` must contain an
`admin9_Hexagon` layer built by `make_hex_grid()` and enriched by
`make_admin_lookup()`.

```{r}
#| eval: false
library(devPTIpack)

my_shp    <- readRDS("app-data/shapes.rds")
hex_layer <- my_shp$admin9_Hexagon
hex_ids   <- hex_layer$admin9Pcod
```

## A. Discover available variables {#a-discover}

Browse the full registry catalogue:

```{r}
#| eval: false
list_hex_vars()
```

Returns a tibble — one row per registered variable — with columns:
`source_id`, `source_label`, `canonical_name`, `var_name`, `var_units`,
`time_col`, `available_years`, `weight`, `fun`, `is_population`.

To query the **live** parquet for the actual years present (not just the
registry hint):

```{r}
#| eval: false
get_available_years("flood_exposure_15cm_1in100")
#> flood_exposure_15cm_1in100 (wb_flood_exposure): 2020
```

Pass a single canonical name (character) or a `pti_hex_var` object
returned by `use_hex_vars()`.

## B. Configure variable selection {#b-configure}

```{r}
#| eval: false
vars <- use_hex_vars(
  "flood_exposure_15cm_1in100",
  years = c(2020)
)
```

Population is **always fetched automatically** from the registry —
do not list it here.

**Year resolution rules (applied per variable):**

| Situation | Outcome |
|---|---|
| Exact year available | Used as-is, no message. |
| Exact year absent; one nearer alternative within 7 years | Nearest year substituted; later year on equidistant tie. A warning names the substitution. |
| No data within 7 years of the requested year | Error; available years listed. |

With `years = NULL` in an **interactive session**, the system prints
available years and prompts you to choose. In a **non-interactive
session** (`quarto render`, CI), `years = NULL` errors with an
actionable message — always supply explicit years when scripting.

## C. Fetch hex data {#c-fetch}

```{r}
#| eval: false
hex_data <- fetch_hex_data(hex_ids = hex_ids, vars = vars)
```

Only hexagons matching `hex_ids` are read (parquet predicate pushdown).
Temporal variables are pivoted wide:
`flood_exposure_15cm_1in100_2020`, etc.
Population is always the first indicator column after `hex_id`.

**Resolution bridge:**

| Your grid resolution | Source data | Behaviour |
|---|---|---|
| H6 (same as source) | H6 | Direct fetch — no conversion. |
| H5 (coarser) | H6 | Each H5 cell expanded to its ~7 H6 children; fetched at H6; aggregated back to H5 before returning. |
| H7 (finer than source) | H6 | Error — finer than source is not supported. |

## C′. Optional: merge your own local parquet {#c-prime-merge}

Merge project-specific indicators between fetch and aggregate:

```{r}
#| eval: false
my_data <- arrow::open_dataset("data-raw/conflict.parquet") |>
  dplyr::filter(h3id %in% hex_ids) |>
  dplyr::collect() |>
  dplyr::rename(hex_id = h3id) |>
  tidyr::pivot_wider(
    names_from  = year,
    values_from = conflict_events,
    names_prefix = "conflict_"
  )

hex_data <- dplyr::full_join(hex_data, my_data, by = "hex_id")
```

**Local parquet contract:**

| Requirement | Rule |
|---|---|
| Format | Readable by `arrow::open_dataset()` |
| Layout | Long format — one row per hex × time period |
| Hex column | Any name — rename to `hex_id` before the `full_join` |
| Population | Absent — comes from the registry; do not duplicate it |
| Values | Numeric columns, one per indicator |
| Uniqueness | No duplicate hex ID per time period |

Custom variables are not in the registry, so you **must** supply an
`indicator_config` tibble in step E for each one.

## D. Aggregate to admin polygons {#d-aggregate}

```{r}
#| eval: false
aggregated <- aggregate_hex_to_shapes(
  hex_data  = hex_data,
  hex_layer = hex_layer,
  shp_dta   = my_shp,
  strategy  = list(
    .default               = c(weight = "pop",  fun = "mean"),
    flood_exposure_15cm_1in100 = c(weight = "pop", fun = "mean"),
    conflict               = c(weight = "none", fun = "sum")
  )
)
```

**Strategy has two orthogonal dimensions:**

- **`weight`**: `"pop"` (population-weighted), `"area"` (area-weighted),
  or `"none"` (unweighted).
- **`fun`**: `"mean"`, `"median"`, `"sum"`, `"min"`, `"max"`.

A `.default` entry covers all variables not explicitly named.  For
temporal variables, use the **stem** name without the year suffix
(e.g. `flood_exposure_15cm_1in100` covers
`flood_exposure_15cm_1in100_2020` and any other resolved years).

Aggregation is always **hex → each admin level directly**
(hex→admin1, hex→admin2, etc.) — never chained through intermediate
levels.  `NA` values in individual hexes are treated as missing and
excluded from the weighted calculation; if all hexes in a polygon are
`NA` for a variable, the polygon result is `NA` with a warning.

## E. Build metadata Excel {#e-build-metadata-excel}

```{r}
#| eval: false
build_hex_metadata(
  aggregated       = aggregated,
  shp_dta          = my_shp,
  # indicator_config needed only for non-registry variables (e.g. conflict):
  indicator_config = tibble::tibble(
    canonical_name        = "conflict",
    var_name              = "Conflict Events ({year})",
    var_description       = "ACLED conflict event count per admin polygon.",
    var_units             = "count",
    pillar_group          = 4,
    pillar_name           = "Security",
    legend_revert_colours = TRUE,
    fltr_exclude_pti      = FALSE
  ),
  country_name       = "Rwanda",
  output_path        = "app-data/metadata-hex.xlsx",
  include_hex        = INCLUDE_HEX_IN_APP,
  # include_population = FALSE  # default: population excluded from metadata
  #                             # (avoids collision with Step 3 population).
  #                             # Set to TRUE if you want population as a
  #                             # visible map indicator in the deployed app.
)
```

**Registry auto-population:** variables declared in the bundled
`inst/hex_vars_registry.yaml` (e.g. `flood_exposure_15cm_1in100`)
have their `var_name`, `var_units`, `pillar_name`, `legend_revert_colours`,
and `fltr_exclude_pti` pre-filled from the registry.  Pass
`indicator_config` only for non-registry indicators, or to **override**
a registry default — a warning is emitted when a registry value is
overridden.

**Population exclusion:** by default (`include_population = FALSE`),
population is used internally for aggregation weighting but is **not**
written to the metadata sheet.  This avoids a duplicate `population`
row when Step 3 already provides population statistics.  Set
`include_population = TRUE` if you want population as a separate
visible indicator in the app (e.g. raw hex-derived population counts).

**Temporal variables** expand to one metadata row per resolved year, with
`var_name` generated from the glue template — `{year}` is replaced with
the actual year (e.g. `"Conflict Events (2020)"`).

### The `INCLUDE_HEX_IN_APP` switch

Set in `00-master.R`:

```r
INCLUDE_HEX_IN_APP <- FALSE
```

| Value | Effect on `metadata-hex.xlsx` | Effect on the deployed app |
|---|---|---|
| `FALSE` (default) | Hex sheet omitted. Only aggregated admin-level sheets written. | Smaller app — hex polygons not shipped. Indicator values still appear at every admin level. |
| `TRUE` | `admin9_Hexagon` sheet included. | Hex polygons shipped. Users can explore the raw hex grid in the app. |

When `include_hex` is not supplied and the hex grid exceeds 5,000 cells,
the function prompts interactively (and defaults to `FALSE`
in non-interactive sessions with a warning).

### Review `fltr_exclude_pti` before Step 5

::: {.callout-warning}
**Before running Step 5:** open `app-data/metadata-hex.xlsx` and review
the `fltr_exclude_pti` column for each variable.  Registry variables
marked `fltr_exclude_pti = TRUE` (e.g. `population`) are excluded from
the PTI weight sliders but still visible as overlays.  Non-registry
variables default to `FALSE` (included in PTI).  Adjust any values
that do not match your project's analytical design before compiling.
:::

## Supported admin levels

Hex-derived indicators appear on **every** admin-level sheet in the
output Excel (admin1, admin2, etc.).  The `spatial_level` column in the
metadata sheet points to the **hex level** (e.g. `admin9_Hexagon`) —
the finest available resolution, regardless of which admin-level sheet
the user is viewing.

## Next

Once `app-data/metadata-hex.xlsx` is produced, continue with
[Step 5 — Compile & finalise](build-pti-5-compile.html).
